Efficient High-Level Abstractions for Web Programming

### 2013-09-02 papers/gpce2013/gpce2013.tex
Add info on the size of the Chooze application
Capitalization

We classified existing approaches providing high-level abstractions for Web programming in four
categories, as shown in figure \ref{fig:langs}.

@@

Another approach consists in defining several independent domain-specific
languages~\cite{Van00_DSL}, each one focusing on concerns specific to a given problem domain, and
then to combine all the source artifacts written with these language into one executable program, as
shown in figure~\ref{fig:langs} (b). Defining such languages requires a minimal effort compared to
the previous approach because each language has a limited set of features. On the other hand, it is
difficult to have interoperability between DSLs. \cite{Visser07} gave an example of such
a domain-specific language for defining Web applications.

@@

%\vspace{0.1cm}
The last approach, shown in figure \ref{fig:langs} (d), can be seen as a middle-ground between the
two previous approaches: DSLs are embedded in a host language but use a code generation process.
This approach shares the same benefits and limitations as embedded DSLs for defining language units.
However, the code generation process is specific to each DSL and gives the opportunity to perform
domain-specific optimizations. In other words deeply embedded DSLs bring domain-specific knowledge
to the compiler. Js-scala~\cite{Kossakowski12_JsDESL} is an example of deeply embedded DSL in Scala
for Web programming. It makes it possible to produce JavaScript programs from Scala code that uses
basic language concepts like arrays and control structures (\code{if} and \code{while}) as well as
mechanisms specific to the Scala compiler, like delimited continuations to handle asynchronous
computations. Though the initial authors of js-scala showed how to use the native browser APIs in an
untyped fashion, they did not show how to build high-level abstractions for Web programming
generating efficient code leveraging the browser APIs and did not show how to share such
abstractions between server and client sides (they achieved code sharing, yet on the server-side
their DSL was not compiled but interpreted). Our paper addresses these limitations.

@@

A high-level abstraction for searching elements in a document could be just one function
finding all elements matching a given CSS selector. In fact, most JavaScript
developers\footnote{According to
\href{http://trends.builtwith.com/javascript}{http://trends.builtwith.com/javascript}, jQuery is
used by more than 40\% of the top million sites.} use the jQuery library~\cite{Bibeault08_jQuery}
that actually provides only one function to search for elements. Listing \ref{jquery-selectors}
shows an equivalent JavaScript program as listing \ref{vanilla-selectors}, but using jQuery. The
code is both shorter and simpler, thanks to its higher level of abstraction. jQuery provides an API
that is simpler to master because it has fewer functions, but this benefit comes at the price of a
decrease in runtime performance.

@@

Finally, listing~\ref{js-scala-selectors} shows how to implement listing~\ref{jquery-selectors} in
Scala using js-scala. The code has the same abstraction level as with jQuery, however it
generates a JavaScript program identical to listing~\ref{vanilla-selectors}: the high-level
abstractions (the \code{find} and \code{findAll} functions) exist only in the initial program, not
in the final JavaScript program.

@@

\code{null} references are a known source of problems in programming
languages~\cite{Hoare09_Null,Nanda09_Null}. For example, consider listing \ref{null-unsafe} finding
a particular widget in the page and then a particular button within the widget. The native
\code{querySelector} method returns \code{null} if no node matched the given selector in the
document. If we run this code in a page where the widget is not present, it will throw an error
and stop further JavaScript execution. Defensive code can be written to handle \code{null}
references, but leads to very cumbersome code, as shown in listing \ref{null-defensive}.

@@

In our DSL, we encode an optional value of type \code{Rep[A]} using a \code{Rep[Option[A]]} value,
which can either be a \code{Rep[Some[A]]} (if there is a value) or a \code{Rep[None.type]} (if
there is no value). An optional value can be dereferenced using the \code{for} notation, as shown
in listing \ref{null-js-scala}, that implements in js-scala a program equivalent to listing
\ref{null-defensive}. The \code{find} function returns a \code{Rep[Option[Element]]}. The \code{for}
expression contains a sequence of statements that are executed in order, as long as the previous
statement returned a \code{Rep[Some[Element]]} value.

@@

Such a monadic API brings both safety and expressiveness to developers manipulating optional values
but usually involves the creation of an extra container object holding the optional value. In our
case, the monadic API is used in the initial program but generates code that does not wrap values in
container objects but instead checks if they are \code{null} or not when dereferenced. So the extra
container object exists only in the initial program and is removed during code generation: listing
\ref{null-js-scala} produces a code equivalent to listing~\ref{null-defensive}.

@@

Listing \ref{option-codegen} shows the JavaScript code generator for methods \code{isEmpty} (that
checks if the optional value contains a value) and \code{foreach} (that is called when the
\code{for} notation is used, as in listing \ref{null-js-scala}). The \code{emitNode} method handles
\code{OptionIsEmpty} and \code{OptionForeach} nodes returned by the implementations of
\code{isEmpty} and \code{foreach}, respectively. In the case of the \code{OptionIsEmpty} node, it
simply generates an expression testing if the value is \code{null}. In the case of the
\code{OptionForeach} node, it wraps the code block dereferencing the value within a \code{if}
checking that the value is not \code{null}.

@@

A common task in Web applications consists in computing HTML fragments representing a part of the
page content. This task can be performed either from the server-side (to initially respond to a
request) or from the client-side (to update the current page). As an example, listing \ref{dom-api}
defines a JavaScript function \code{articleUi} that builds a DOM tree containing an article
description, and listing \ref{scala-xml-api} shows how one could implement a similar function on
server-side using the standard Scala XML library. The reader may notice that the client-side
and server-side APIs are very different and that the client-side API is very low-level and
inconvenient to use.

Our first step consists in capturing, in a high-level API, the concepts common to the JavaScript and
Scala APIs. Though they are different, both APIs define HTML elements with attributes and content.
We propose to have a function \code{el} to define an HTML element, eventually containing attributes
and children elements. Any children of an element that is not an element itself is converted into a
text node. Listing \ref{forest} shows how to implement our example with our DSL. The children
elements of an element can also be obtained dynamically from a collection, as shown in listing
\ref{forest-loops}.

@@

The \code{el} function returns an \code{Element} IR node that is a tree composed of other
\code{Element} and \code{Text} nodes. The JavaScript and Scala code generators traverse this tree
and produce code building an equivalent DOM tree and XML fragment, respectively. When the children
of an element are constant values (as in listing \ref{forest}) rather than dynamically computed (as
in listing \ref{forest-loops}), the code generators unroll the loop that adds children to their
parent, for better performance. As a result, listing \ref{forest} generates a code equivalent to
listing \ref{dom-api} on client-side and equivalent to listing \ref{scala-xml-api} on server-side.

@@

We could not implement this abstraction in GWT or HaXe as efficiently as we did in js-scala because
it relies on the staging mechanism: the best we could do in GWT or Haxe is to expose the native
high-level API (\code{querySelector} and \code{querySelectorAll}). So we directly compared the
execution time of the JavaScript code generated by listing \ref{js-scala-selectors} with a
JavaScript program equivalent to listing \ref{vanilla-selectors} but using the high-level native
API (\code{querySelector} and \code{querySelectorAll}) instead. The code was executed in a Web page
containing a few elements: 4 \code{fieldset} elements, each containing 0 to 2 elements with class
\code{word}.

@@

Figure \ref{fig:selectors-benchmark} shows the benchmark results. The JavaScript-opt version is
listing \ref{vanilla-selectors}, which uses low-level native APIs, the JavaScript version is the
equivalent listing using the high-level native API, and the jQuery version is listing
\ref{jquery-selectors}. The js-scala version is slightly slower than the JavaScript-opt (by 14\%),
but is 2.88 times faster than the JavaScript version, and 28.6 times faster than the jQuery
version. Finally, the js-scala version has a performance to lines of code ratio more than 1.72
times higher than others.

@@

Chooze~\footnote{Source code is available at
\href{http://github.com/julienrf/chooze}{http://github.com/julienrf/chooze}} is an existing
complete application for making polls. It allows users to create a poll, define the choice
alternatives, share the poll, vote and look at the results. It contains JavaScript code to handle
the dynamic behavior of the application: double-posting prevention, dynamic form update and rich
interaction with the document.

>>>>

We classified existing approaches providing high-level abstractions for Web programming in four
categories, as shown in Figure \ref{fig:langs}.

@@

Another approach consists in defining several independent domain-specific
languages~\cite{Van00_DSL}, each one focusing on concerns specific to a given problem domain, and
then to combine all the source artifacts written with these language into one executable program, as
shown in Figure~\ref{fig:langs} (b). Defining such languages requires a minimal effort compared to
the previous approach because each language has a limited set of features. On the other hand, it is
difficult to have interoperability between DSLs. \cite{Visser07} gave an example of such
a domain-specific language for defining Web applications.

@@

%\vspace{0.1cm}
The last approach, shown in Figure \ref{fig:langs} (d), can be seen as a middle-ground between the
two previous approaches: DSLs are embedded in a host language but use a code generation process.
This approach shares the same benefits and limitations as embedded DSLs for defining language units.
However, the code generation process is specific to each DSL and gives the opportunity to perform
domain-specific optimizations. In other words deeply embedded DSLs bring domain-specific knowledge
to the compiler. Js-scala~\cite{Kossakowski12_JsDESL} is an example of deeply embedded DSL in Scala
for Web programming. It makes it possible to produce JavaScript programs from Scala code that uses
basic language concepts like arrays and control structures (\code{if} and \code{while}) as well as
mechanisms specific to the Scala compiler, like delimited continuations to handle asynchronous
computations. Though the initial authors of js-scala showed how to use the native browser APIs in an
untyped fashion, they did not show how to build high-level abstractions for Web programming
generating efficient code leveraging the browser APIs and did not show how to share such
abstractions between server and client sides (they achieved code sharing, yet on the server-side
their DSL was not compiled but interpreted). Our paper addresses these limitations.

@@

A high-level abstraction for searching elements in a document could be just one function
finding all elements matching a given CSS selector. In fact, most JavaScript
developers\footnote{According to
\href{http://trends.builtwith.com/javascript}{http://trends.builtwith.com/javascript}, jQuery is
used by more than 40\% of the top million sites.} use the jQuery library~\cite{Bibeault08_jQuery}
that actually provides only one function to search for elements. Listing \ref{jquery-selectors}
shows an equivalent JavaScript program as Listing \ref{vanilla-selectors}, but using jQuery. The
code is both shorter and simpler, thanks to its higher level of abstraction. jQuery provides an API
that is simpler to master because it has fewer functions, but this benefit comes at the price of a
decrease in runtime performance.

@@

Finally, Listing~\ref{js-scala-selectors} shows how to implement Listing~\ref{jquery-selectors} in
Scala using js-scala. The code has the same abstraction level as with jQuery, however it
generates a JavaScript program identical to Listing~\ref{vanilla-selectors}: the high-level
abstractions (the \code{find} and \code{findAll} functions) exist only in the initial program, not
in the final JavaScript program.

@@

\code{null} references are a known source of problems in programming
languages~\cite{Hoare09_Null,Nanda09_Null}. For example, consider Listing \ref{null-unsafe} finding
a particular widget in the page and then a particular button within the widget. The native
\code{querySelector} method returns \code{null} if no node matched the given selector in the
document. If we run this code in a page where the widget is not present, it will throw an error
and stop further JavaScript execution. Defensive code can be written to handle \code{null}
references, but leads to very cumbersome code, as shown in Listing \ref{null-defensive}.

@@

In our DSL, we encode an optional value of type \code{Rep[A]} using a \code{Rep[Option[A]]} value,
which can either be a \code{Rep[Some[A]]} (if there is a value) or a \code{Rep[None.type]} (if
there is no value). An optional value can be dereferenced using the \code{for} notation, as shown
in Listing \ref{null-js-scala}, that implements in js-scala a program equivalent to Listing
\ref{null-defensive}. The \code{find} function returns a \code{Rep[Option[Element]]}. The \code{for}
expression contains a sequence of statements that are executed in order, as long as the previous
statement returned a \code{Rep[Some[Element]]} value.

@@

Such a monadic API brings both safety and expressiveness to developers manipulating optional values
but usually involves the creation of an extra container object holding the optional value. In our
case, the monadic API is used in the initial program but generates code that does not wrap values in
container objects but instead checks if they are \code{null} or not when dereferenced. So the extra
container object exists only in the initial program and is removed during code generation: Listing
\ref{null-js-scala} produces a code equivalent to Listing~\ref{null-defensive}.

@@

Listing \ref{option-codegen} shows the JavaScript code generator for methods \code{isEmpty} (that
checks if the optional value contains a value) and \code{foreach} (that is called when the
\code{for} notation is used, as in Listing \ref{null-js-scala}). The \code{emitNode} method handles
\code{OptionIsEmpty} and \code{OptionForeach} nodes returned by the implementations of
\code{isEmpty} and \code{foreach}, respectively. In the case of the \code{OptionIsEmpty} node, it
simply generates an expression testing if the value is \code{null}. In the case of the
\code{OptionForeach} node, it wraps the code block dereferencing the value within a \code{if}
checking that the value is not \code{null}.

@@

A common task in Web applications consists in computing HTML fragments representing a part of the
page content. This task can be performed either from the server-side (to initially respond to a
request) or from the client-side (to update the current page). As an example, Listing \ref{dom-api}
defines a JavaScript function \code{articleUi} that builds a DOM tree containing an article
description, and Listing \ref{scala-xml-api} shows how one could implement a similar function on
server-side using the standard Scala XML library. The reader may notice that the client-side
and server-side APIs are very different and that the client-side API is very low-level and
inconvenient to use.

Our first step consists in capturing, in a high-level API, the concepts common to the JavaScript and
Scala APIs. Though they are different, both APIs define HTML elements with attributes and content.
We propose to have a function \code{el} to define an HTML element, eventually containing attributes
and children elements. Any children of an element that is not an element itself is converted into a
text node. Listing \ref{forest} shows how to implement our example with our DSL. The children
elements of an element can also be obtained dynamically from a collection, as shown in Listing
\ref{forest-loops}.

@@

The \code{el} function returns an \code{Element} IR node that is a tree composed of other
\code{Element} and \code{Text} nodes. The JavaScript and Scala code generators traverse this tree
and produce code building an equivalent DOM tree and XML fragment, respectively. When the children
of an element are constant values (as in Listing \ref{forest}) rather than dynamically computed (as
in Listing \ref{forest-loops}), the code generators unroll the loop that adds children to their
parent, for better performance. As a result, Listing \ref{forest} generates a code equivalent to
Listing \ref{dom-api} on client-side and equivalent to Listing \ref{scala-xml-api} on server-side.

@@

We could not implement this abstraction in GWT or HaXe as efficiently as we did in js-scala because
it relies on the staging mechanism: the best we could do in GWT or Haxe is to expose the native
high-level API (\code{querySelector} and \code{querySelectorAll}). So we directly compared the
execution time of the JavaScript code generated by Listing \ref{js-scala-selectors} with a
JavaScript program equivalent to Listing \ref{vanilla-selectors} but using the high-level native
API (\code{querySelector} and \code{querySelectorAll}) instead. The code was executed in a Web page
containing a few elements: 4 \code{fieldset} elements, each containing 0 to 2 elements with class
\code{word}.

@@

Figure \ref{fig:selectors-benchmark} shows the benchmark results. The JavaScript-opt version is
Listing \ref{vanilla-selectors}, which uses low-level native APIs, the JavaScript version is the
equivalent listing using the high-level native API, and the jQuery version is Listing
\ref{jquery-selectors}. The js-scala version is slightly slower than the JavaScript-opt (by 14\%),
but is 2.88 times faster than the JavaScript version, and 28.6 times faster than the jQuery
version. Finally, the js-scala version has a performance to lines of code ratio more than 1.72
times higher than others.

@@

Chooze~\footnote{Source code is available at
\href{http://github.com/julienrf/chooze}{http://github.com/julienrf/chooze}} is an existing
complete application for making polls. It allows users to create a poll, define the choice
alternatives, share the poll, vote and look at the results. It contains JavaScript code to handle
the dynamic behavior of the application: double-posting prevention, dynamic form update and rich
interaction with the document. The size of the whole application (server and client sides) is about one thousand of lines of code.
